---
title: "Using Rust types in R"
---

The `#[extendr]` attribute macro is how Rust types are made available to R.
This chapter covers how to use it on structs, enums, and impl blocks.

::: {.callout-note}
If you would like to learn more about Rust macros, check The Book,
specifically
[Part 5 of Ch. 20](https://doc.rust-lang.org/book/ch20-05-macros.html). You
may also want to consult the section on
[procedural macros](https://doc.rust-lang.org/reference/procedural-macros.html)
in the official Rust reference manual.
:::

```{r}
#| include: false
library(rextendr)
```

## Exporting `impl` blocks

The `#[extendr]` macro works with inherent implementations on a type such as
an `enum` or a `struct`. extendr *does not* support using `#[extendr]` on
trait impls.

::: callout-note
You can only add an inherent implementation on a type that you own and not one
provided by a third-party crate. Doing so would violate the [orphan
rules](https://github.com/Ixrec/rust-orphan-rules?tab=readme-ov-file#what-are-the-orphan-rules).
See [Wrapping third-party types](#wrapping-third-party-types) for how to
handle that case.
:::

Without `#[extendr]` on the type itself, trying to return it from a function
results in a compilation error:

```{.rust}
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
fn make_shape(shape: &str) -> Shape {
    match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!()
    }
}
```
```
error[E0277]: the trait bound `Shape: ToVectorValue` is not satisfied
  --> src/lib.rs:11:1
   |
11 | #[extendr]
   | ^^^^^^^^^^ unsatisfied trait bound
   |
help: the trait `ToVectorValue` is not implemented for `Shape`
  --> src/lib.rs:4:1
   |
 4 | enum Shape {
   | ^^^^^^^^^^
   = help: the following other types implement trait `ToVectorValue`:
             &&str
             &(f64, f64)
             &Rbool
             &Rcplx
             &Rfloat
             &Rint
             &String
             &bool
           and 48 others
   = note: required for `extendr_api::Robj` to implement `From<Shape>`
   = note: this error originates in the attribute macro `extendr` (in Nightly builds, run with -Z macro-backtrace for more info)
```

Adding `#[extendr]` to the enum itself generates the necessary `From`/`TryFrom`
implementations between the type and `Robj`, making it returnable to R:

```{extendrsrc}
#| label: shape-enum-no-impl
#[derive(Debug)]
#[extendr]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
fn make_shape(shape: &str) -> Shape {
    match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!()
    }
}
```

It is also possible to add methods to structs/enums and their instances:

```{extendrsrc}
#| label: shape-enum-with-impl
#[derive(Debug)]
#[extendr]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
impl Shape {
    fn new(x: &str) -> Self {
        match x {
            "triangle" => Self::Triangle,
            "rectangle" => Self::Rectangle,
            "pentagon" => Self::Pentagon,
            "hexagon" => Self::Hexagon,
            &_ => unimplemented!(),
        }
    }

    fn n_coords(&self) -> usize {
        match &self {
            Shape::Triangle => 3,
            Shape::Rectangle => 4,
            Shape::Pentagon => 4,
            Shape::Hexagon => 5,
        }
    }
}
```

In this example two new methods are added to the `Shape` enum. The first,
`new()`, is like the `make_shape()` function that was shown earlier: it takes
a `&str` and returns an enum variant. Now that the enum has an `impl` block
with the `#[extendr]` attribute macro, it can be exported to R by inclusion in
the `extendr_module! {}` macro.

```rust
extendr_module! {
    mod hellorust;
    impl Shape;
}
```

::: callout-important
You can only have **one** `#[extendr]` impl block per type per module.
:::

Doing so creates an environment in your package called `Shape`. The environment
contains all of the methods that are available to you.

If you run `as.list(Shape)` you will see that there are two functions in the
environment which enable you to call the methods defined in the impl block. You
might think that this feels like an [R6
object](https://r6.r-lib.org/articles/Introduction.html) and you'd be right
because an R6 object essentially is an environment!

```{r}
as.list(Shape)
```

Calling the `new()` method instantiates a new enum variant.

```{r}
tri <- Shape$new("triangle")
tri
```

The newly made `tri` object is an [external
pointer](https://cran.r-project.org/doc/manuals/R-exts.html#External-pointers-and-weak-references)
to the `Shape` enum in Rust. This pointer has the same methods as the Shape
environment—though they cannot be seen in the same way. For example you can
run the `n_coords()` method on the newly created object.

```{r}
tri$n_coords()
```

::: callout-tip
To make the methods visible to the `Shape` class you can define a
`.DollarNames` method which will allow you to preview the methods and
attributes when using the `$` syntax. This is very handy to define when making
an impl a core part of your package.

```{r}
.DollarNames.Shape = function(env, pattern = "") {
  ls(Shape, pattern = pattern)
}
```
:::

### `impl` ownership

Adding the `#[extendr]` macro to an impl allows the struct or enum to be made
available to R as an external pointer. Once you create an external pointer,
that is then owned by R. So you can only get references to it or mutable
references. If you need an owned version of the type, then you will need to
clone it.

## Accessing exported `impl`s from Rust

Invariably, if you have made an impl available to R via the `#[extendr]` macro,
you may want to define functions that take the impl as a function argument.

Due to R owning the `impl`'s external pointer, these functions cannot take an
owned version of the impl as an input. For example, trying to define a function
that subtracts an integer from the `n_coords()` output like below returns a
compiler error:

```{.rust}
#[extendr]
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
impl Shape {
    fn new(x: &str) -> Self {
        match x {
            "triangle" => Self::Triangle,
            "rectangle" => Self::Rectangle,
            "pentagon" => Self::Pentagon,
            "hexagon" => Self::Hexagon,
            &_ => unimplemented!(),
        }
    }

    fn n_coords(&self) -> usize {
        match &self {
            Shape::Triangle => 3,
            Shape::Rectangle => 4,
            Shape::Pentagon => 4,
            Shape::Hexagon => 5,
        }
    }
}

#[extendr]
fn subtract_coord(x: Shape, n: i32) -> i32 {
    (x.n_coords() as i32) - n
}
```
```
error[E0277]: the trait bound `Shape: TryFrom<...>` is not satisfied
  --> src/lib.rs:33:1
   |
33 | #[extendr]
   | ^^^^^^^^^^ unsatisfied trait bound
   |
help: the trait `From<&extendr_api::Robj>` is not implemented for `Shape`
  --> src/lib.rs:4:1
   |
 4 | enum Shape {
   | ^^^^^^^^^^
   = note: required for `&extendr_api::Robj` to implement `Into<Shape>`
   = note: required for `Shape` to implement `TryFrom<&Robj>`
   = note: required for `&extendr_api::Robj` to implement `TryInto<Shape>`
   = note: the full name for the type has been written to 
```

In this case, we can take a reference to `&Shape` instead since the struct is owned by R now.

## Wrapping third-party types

Sometimes you want to expose a type from a third-party crate to R. Because you
don't own that type, you cannot add `#[extendr]` directly to it—doing so would
violate the orphan rules. The solution is a *newtype wrapper*: a struct you own
that contains the foreign type as its only field.

Applying `#[extendr]` to the wrapper struct generates the `From`/`TryFrom`
implementations between the wrapper and `Robj`, and an `#[extendr]` impl block
then exposes methods just as with any type you own. The
[`tomledit`](https://github.com/extendr/tomledit/blob/main/src/rust/src/lib.rs)
package is a good real-world example: it wraps `toml_edit::DocumentMut` in a
`Toml` newtype and exposes parsing, reading, and writing methods to R.

As a simplified illustration:

```{extendrsrc}
#| label: newtype-wrapper
// Pretend `Inner` comes from a third-party crate we don't own.
#[derive(Debug, Clone)]
struct Inner(i32);

// Our newtype wrapper — we own this, so #[extendr] is allowed.
#[extendr]
#[derive(Debug, Clone)]
struct Wrapper(Inner);

#[extendr]
impl Wrapper {
    fn new(x: i32) -> Self {
        Wrapper(Inner(x))
    }

    fn value(&self) -> i32 {
        self.0.0
    }
}
```

```{r}
w <- Wrapper$new(42L)
w$value()
```

::: callout-tip
Deriving `Clone` on your newtype wrapper is often necessary. Because R owns
the external pointer, methods that produce a modified copy must clone the inner
value and return the new wrapper rather than mutating in place.
:::

## Documenting methods

Methods on `#[extendr]` impl blocks can be documented with Roxygen-style `///`
doc comments. These are picked up by `rextendr` when generating R wrappers,
so the documentation surfaces in R's help system via `?`. The
[`fiorefactor`](https://github.com/albersonmiranda/fiorefactor/blob/main/src/rust/src/leontief.rs)
package is a good example of this in practice.

The `///` prefix is Rust's item-level doc comment syntax. Tags follow standard
Roxygen2 conventions: `@param`, `@return`, `@details`, `@examples`, and so on.

```{extendrsrc}
#| label: doc-comments
#[extendr]
#[derive(Debug, Clone)]
struct Counter(i32);

#[extendr]
impl Counter {
    /// Create a new Counter starting at zero.
    /// @return A `Counter` object.
    /// @export
    fn new() -> Self {
        Counter(0)
    }

    /// Increment the counter by one.
    /// @return Self (invisibly)
    /// @export
    fn increment(&mut self) {
        self.0 += 1;
    }

    /// Return the current count.
    /// @return An integer.
    /// @export
    fn value(&self) -> i32 {
        self.0
    }
}
```

```{r}
c1 <- Counter$new()
c1$increment()
c1$increment()
c1$value()
```
