---
title: "Introduction"
description: "Call Rust code in R!"
weight: 2
---


```{r}
#| label: set-root-dir
#| include: false
root_dir <- tempdir()
knitr::opts_knit$set(root.dir = root_dir)
```

This is the website for `extendr`, a developer-first utility for building R
packages that call Rust code. Of course, you do not have to be a package 
developer to find some use in extendr. In fact, anyone who has a need to call 
Rust in R can do so with ease. Consider this basic example:

```{r}
#| label: rust-function
# create a Rust function
rextendr::rust_function(
  "fn add(a:f64, b:f64) -> f64 { a + b }",
  quiet = TRUE
)

# call it from R
add(2.5, 4.7)
```

Moving beyond these simple, one-off cases, however, is where extendr really 
shines. Any time you have lots of really complex Rust code that requires 
seamless integration with R, notably within a package, extendr is the tool for
you. 

Now largely owing to the fact that we are bridging programming languages, you 
will rarely proceed directly to extendr's Rust API. You will instead reach for 
the R package `rextendr`, which works sort of like `usethis`, automating many of 
the repetitive tasks required to setup or scaffold your project, so you can 
focus on the important business of writing code. This is done by calling
`rextendr::use_extendr()`, which should feel familiar to anyone who has worked
with `usethis::use_cpp11()`.

## Create an R package

The process of building a Rust-powered R package with extendr boils down to 
three simple steps:

1. Setup R package with `usethis::create_package()`.
2. Add extendr scaffolding with `rextendr::use_extendr()`.
3. Build and document with `devtools::document()`.

Once you have worked through those steps, you can then load your package with 
`devtools::load_all()` and start using your package! Here is an example with 
just the defaults:

```r
# setup R package
usethis::create_package("helloextendr")

# add extendr scaffolding
rextendr::use_extendr()

# build and document package
devtools::document()

# load package
devtools::load_all()

# call rust function
hello_world()
```

```
[1] "Hello world!"
```

A Rust-powered R package in one fell swoop! Now, if you want to see a complete 
example of an actual R package, check out the [heck example](@/example.md).

## What's next?

-   If you do not already have Rust, R, and rextendr, check out our [Installation guide](@/installation.md).
-   Want to learn more about Rust? Check out our [Rust tutorials](@/rust-intro/_index.md).
-   Ready to dive in to package development? You will want to consult our
    [Development guide](@/development/_index.md).
-   Have a Rust-powered package and want to publish it to CRAN? Jump to our 
    [Publishing recommendations](@/publishing/_index.md).
-   Here to join the extendr team and contribute to new features and 
    maintenance? Have a look at our [Guide for contributors](@/contributing/_index.md).