---
title: "The `#[extendr]` macro"
execute: 
  eval: false
---

The central task of extendr is to make Rust code available in R. Crucial to this
effort is the attribute macro, `#[extendr]`. This chapter provides a brief 
introduction to the use of that macro. 

::: {.callout-note}
If you would like to learn more about Rust macros, check The Book, specifically 
[Part 5 of Ch. 20](https://doc.rust-lang.org/book/ch20-05-macros.html). You may 
also want to consult the section on [procedural macros](https://doc.rust-lang.org/reference/procedural-macros.html)
in the official Rust reference manual.
:::

```{r}
#| include: false
library(rextendr)
```

Rust allows attribute macros to be placed on functions, struct, enums, and 
impls, and extendr supports all of these. The general pattern is mostly the same
in each case, but for starters, let's consider how to apply the extendr macro to
a Rust function in order to make it available in R. In order for that to happen, 
three things need to happen. First, you need to declare the extendr namespace to 
the compiler with `use`. After that, you need to annotate whatever function you 
want to make available in R with the `#[extendr]` macro. Finally, you must 
register the function in the `extendr_module! {}` macro. 

As an example, suppose that you have created a function that gives the answer to 
the question "What is the meaning of life?" and that you have very thoughtfully 
decided to share this wisdom with the R community. Your main Rust library script 
should then look like this:

```{.rust filename="src/lib.rs"}
use extendr_api::prelude::*;

#[extendr]
fn answer_to_life() -> i32 {
    42
}

extendr_module! {
    mod hitchhiker;
    fn answer_to_life;
}
```

Here, the `#[extendr]` macro is placed directly above the function definition.


## `ToVectorValue` trait

In order for an item to be returned from a function marked with the `#[extendr]`
attribute macro, it must be able to be turned into an R object. In extendr, the
struct `Robj` is a catch all for any type of R object.

::: callout-note
For those familiar with PyO3, the `Robj` struct is similar in concept to the
[`PyAny`](https://docs.rs/pyo3/latest/pyo3/types/struct.PyAny.html) struct.
:::

The `ToVectorValue` trait is what is used to convert Rust items into R objects.
The trait is implemented on a number of standard Rust types such as `i32`,
`f64`, `usize`, `String` and more (see [all foreign implementations
here](https://extendr.github.io/extendr/extendr_api/robj/into_robj/trait.ToVectorValue.html#foreign-impls))
which enables these functions to be returned from a Rust function marked with
`#[extendr]`.

::: callout-note
In essence, all items that are returned from a function must be able to be
turned into an `Robj`. Other extendr types such as `List`, for example, have a
`From<T> for Robj` implementation that defines how it is converted into an
`Robj`.
:::

This means that with a little extra work, the `Shape` enum can be returned to R.
To do so, the `#[extendr]` macro needs to be added to an impl block.

## Exporting `impl` blocks

The other supported item that can be made available to R is an
[`impl`](https://doc.rust-lang.org/std/keyword.impl.html) block. `impl` is a
keyword that allows you to *implement* a trait or an inherent implementation.
The `#[extendr]` macro works with inherent implementations. These are `impl`s on
a type such as an `enum` or a `struct`. extendr *does not* support using
`#[extendr]` on trait impls.

::: callout-note
You can only add an inherent implementation on a type that you have own and not
provided by a third party crate. This would violate the [orphan
rules](https://github.com/Ixrec/rust-orphan-rules?tab=readme-ov-file#what-are-the-orphan-rules).
:::

Continuing with the `Shape` example, this enum alone cannot be returned to R.
For example, the following code will result in a compilation error

```rust
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
fn make_shape(shape: &str) -> Shape {
    match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!()
    }
}
```

```         
error[E0277]: the trait bound `Shape: ToVectorValue` is not satisfied
  --> src/lib.rs:19:1
   |
19 | #[extendr]
   | ^^^^^^^^^^ the trait `ToVectorValue` is not implemented for `Shape`, which is required by `extendr_api::Robj: From<Shape>`
   |
```

However, you add the `#[extendr]` attribute to the `Shape` enum, it can be
returned to R.

```{extendrsrc}
#[derive(Debug)]
#[extendr]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
fn make_shape(shape: &str) -> Shape {
    match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!()
    }
}

```

It is also possible to add methods to structs/enums and their instances:

```{extendrsrc}
#[derive(Debug)]
#[extendr]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
impl Shape {
    fn new(x: &str) -> Self {
        match x {
            "triangle" => Self::Triangle,
            "rectangle" => Self::Rectangle,
            "pentagon" => Self::Pentagon,
            "hexagon" => Self::Hexagon,
            &_ => unimplemented!(),
        }
    }

    fn n_coords(&self) -> usize {
        match &self {
            Shape::Triangle => 3,
            Shape::Rectangle => 4,
            Shape::Pentagon => 4,
            Shape::Hexagon => 5,
        }
    }
}
```

In this example two new methods are added to the `Shape` enum. The first `new()`
is like the `make_shape()` function that was shown earlier: it takes a `&str`
and returns an enum variant. Now that the enum has an `impl` block with
`#[extendr]` attribute macro, it can be exported to R by inclusion in the
`extendr_module! {}` macro.

```rust
extendr_module! {
    mod hellorust;
    impl Shape;
}
```

Doing so creates an environment in your package called `Shape`. The environment
contains all of the methods that are available to you.

::: callout-tip
There are use cases where you may not want to expose any methods but do want to
make it possible to return a struct or an enum to the R. You can do this by
adding an empty impl block with the `#[extendr]` attribute macro.
:::

If you run `as.list(Shape)` you will see that there are two functions in the
environment which enable you to call the methods defined in the impl block. You
might think that this feel like an [R6
object](https://r6.r-lib.org/articles/Introduction.html) and you'd be right
because an R6 object essentially is an environment!

```{r}
as.list(Shape)
```

Calling the `new()` method instantiates a new enum variant.

```{r}
tri <- Shape$new("triangle")
tri
```

The newly made `tri` object is an [external
pointer](https://cran.r-project.org/doc/manuals/R-exts.html#External-pointers-and-weak-references)
to the `Shape` enum in Rust. This pointer has the same methods as the Shape
environment—though they cannot be seen in the same way. For example you can run
the `n_coords()` method on the newly created object.

```{r}
tri$n_coords()
```

::: callout-tip
To make the methods visible to the `Shape` class you can define a `.DollarNames`
method which will allow you to preview the methods and attributes when using the
`$` syntax. This is very handy to define when making an impl a core part of your
package.

```{r}
.DollarNames.Shape = function(env, pattern = "") {
  ls(Shape, pattern = pattern)
}
```
:::

### `impl` ownership

Adding the `#[extendr]` macro to an impl allows the struct or enum to be made
available to R as an external pointer. Once you create an external pointer, that
is then owned by R. So you can only get references to it or mutable references.
If you need an owned version of the type, then you will need to clone it.

## Accessing exported `impl`s from Rust

Invariably, if you have made an impl available to R via the `#[extendr]` macro,
you may want to define functions that take the impl as a function argument.

Due to R owning the `impl`'s external pointer, these functions cannot take an
owned version of the impl as an input. For example trying to define a function
that subtracts an integer from the `n_coords()` output like below returns a
compiler error.

```rust
#[extendr]
fn subtract_coord(x: Shape, n: i32) -> i32 {
    (x.n_coords() as i32) - n
}
```

```         
the trait bound `Shape: extendr_api::FromRobj<'_>` is not satisfied
  --> src/lib.rs:53:22
   |
   | fn subtract_coord(x: Shape, n: i32) -> i32 {
   |                      ^^^^^ the trait `extendr_api::FromRobj<'_>` is not implemented for `Shape`
   |
help: consider borrowing here
   |
   | fn subtract_coord(x: &Shape, n: i32) -> i32 {
   |                      +
   | fn subtract_coord(x: &mut Shape, n: i32) -> i32 {
   |                      ++++
```

As most often, the compiler's suggestion is a good one. Use `&Shape` to use a
reference.

## `ExternalPtr`: returning arbitrary Rust types

In the event that you need to return a Rust type to R that doesn't have a
compatible impl or is a type that you don't own, you can use `ExternalPtr<T>`.
The `ExternalPtr` struct allows any item to be captured as a pointer and
returned to R.

Here, for example, an `ExternalPtr<Shape>` is returned from the `shape_ptr()`
function.

::: callout-tip
Anything that is wrapped in `ExternalPtr<T>` must implement the `Debug` trait.
:::

```{extendrsrc}
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
fn shape_ptr(shape: &str) -> ExternalPtr<Shape> {
    let variant = match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!(),
    };

    ExternalPtr::new(variant)
}
```

Using an external pointer, however, is far more limiting than the `impl` block.
For example, you cannot access any of its methods.

```{r, error = TRUE}
tri_ptr <- shape_ptr("triangle")
tri_ptr$n_coords()
```

To use an `ExternalPtr<T>`, you have to go through a bit of extra work for it.

```{extendrsrc, include = FALSE}
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
impl Shape {
    fn new(x: &str) -> Self {
        match x {
            "triangle" => Self::Triangle,
            "rectangle" => Self::Rectangle,
            "pentagon" => Self::Pentagon,
            "hexagon" => Self::Hexagon,
            &_ => unimplemented!(),
        }
    }

    fn n_coords(&self) -> usize {
        match &self {
            Shape::Triangle => 3,
            Shape::Rectangle => 4,
            Shape::Pentagon => 4,
            Shape::Hexagon => 5,
        }
    }
}

#[extendr]
fn shape_ptr(shape: &str) -> ExternalPtr<Shape> {
    let variant = match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!(),
    };

    ExternalPtr::new(variant)
}

#[extendr]
fn n_coords_ptr(x: Robj) -> i32 {
    let shape    = TryInto::<ExternalPtr<Shape>>::try_into(x); 
    
    match shape {
        Ok(shp) => shp.n_coords() as i32,
        Err(_) => 0
    }
}
```

```rust
#[extendr]
fn n_coords_ptr(x: Robj) -> i32 {
    let shape = TryInto::<ExternalPtr<Shape>>::try_into(x); 
    
    match shape {
        Ok(shp) => shp.n_coords() as i32,
        Err(_) => 0
    }
}
```

This function definition takes an `Robj` and from it, tries to create an
`ExternalPtr<Shape>`. Then, if the conversion did not error, it returns the
number of coordinates as an `i32` (R's version of an integer) and if there was
an error converting, it returns 0.

```{r}
tri_ptr <- shape_ptr("triangle")

n_coords_ptr(tri_ptr)

n_coords_ptr(list())
```

For a good example of using `ExternalPtr<T>` within an R package, refer to the
[`b64` R package](https://github.com/extendr/b64).