---
title: "`ExternalPtr`: returning arbitrary Rust types"
---

```{r}
#| include: false
library(rextendr)
```

In the event that you need to return a Rust type to R that doesn't have a
compatible impl or is a type that you don't own, you can use `ExternalPtr<T>`.
The `ExternalPtr` struct allows any item to be captured as a pointer and
returned to R.

Here, for example, an `ExternalPtr<Shape>` is returned from the `shape_ptr()`
function.

::: callout-tip
Anything that is wrapped in `ExternalPtr<T>` must implement the `Debug` trait.
:::

```{extendrsrc}
#| label: external-ptr
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
fn shape_ptr(shape: &str) -> ExternalPtr<Shape> {
    let variant = match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!(),
    };

    ExternalPtr::new(variant)
}
```

Using an external pointer, however, is far more limiting than the `impl` block.
For example, you cannot access any of its methods.

```{r, error = TRUE}
tri_ptr <- shape_ptr("triangle")
tri_ptr$n_coords()
```

To use an `ExternalPtr<T>`, you have to go through a bit of extra work for it.

```{extendrsrc}
#| label: external-ptr-hidden
#| include: false
#[extendr]
#[derive(Debug)]
enum Shape {
    Triangle,
    Rectangle,
    Pentagon,
    Hexagon,
}

#[extendr]
impl Shape {
    fn new(x: &str) -> Self {
        match x {
            "triangle" => Self::Triangle,
            "rectangle" => Self::Rectangle,
            "pentagon" => Self::Pentagon,
            "hexagon" => Self::Hexagon,
            &_ => unimplemented!(),
        }
    }

    fn n_coords(&self) -> usize {
        match &self {
            Shape::Triangle => 3,
            Shape::Rectangle => 4,
            Shape::Pentagon => 4,
            Shape::Hexagon => 5,
        }
    }
}

#[extendr]
fn shape_ptr(shape: &str) -> ExternalPtr<Shape> {
    let variant = match shape {
        "triangle" => Shape::Triangle,
        "rectangle" => Shape::Rectangle,
        "pentagon" => Shape::Pentagon,
        "hexagon" => Shape::Hexagon,
        &_ => unimplemented!(),
    };

    ExternalPtr::new(variant)
}

#[extendr]
fn n_coords_ptr(x: Robj) -> i32 {
    let shape = TryInto::<ExternalPtr<Shape>>::try_into(x);

    match shape {
        Ok(shp) => shp.n_coords() as i32,
        Err(_) => 0
    }
}
```

```rust
#[extendr]
fn n_coords_ptr(x: Robj) -> i32 {
    let shape = TryInto::<ExternalPtr<Shape>>::try_into(x);

    match shape {
        Ok(shp) => shp.n_coords() as i32,
        Err(_) => 0
    }
}
```

This function definition takes an `Robj` and from it, tries to create an
`ExternalPtr<Shape>`. Then, if the conversion did not error, it returns the
number of coordinates as an `i32` (R's version of an integer) and if there was
an error converting, it returns 0.

```{r}
tri_ptr <- shape_ptr("triangle")

n_coords_ptr(tri_ptr)

n_coords_ptr(list())
```

For a good example of using `ExternalPtr<T>` within an R package, refer to the
[`b64` R package](https://github.com/extendr/b64).
